Temperature and wave exposure as drivers of morphology in two kelp species

### 2019-08-02 morphology_paper.Rmd
moved fromtier template. updated shallow data sheet as some data was not being read in. Completed RDAs and output with updated sites and accompying data

```{r software packages, message=TRUE, warning=FALSE, include=FALSE, echo=FALSE}
library(tidyverse)
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(ggord)
library(vegan)
```

@@

```{r wave power map, echo=FALSE, fig.cap='wave power maps'}
library(png)
library(grid)

img_wavepower <- readPNG("figures/CapePen_Total.PNG")
legend_wave_power <- readPNG("figures/CapePen_Legend.PNG")
grid.raster(img_wavepower,legend_wave_power)
```

@@

Significant differences were found between cool temperate and warm temperate regimes for all *L. pallida* morphological characteristics with lamina length, number of digits and thallus mass the exceptions. Certian morphological characteristics such as stipe mass, stipe length, total length and stipe diameter had significanlty higher means kelps in cool temperate regimes when compared to kelps for the warm temperate regime. Conversely, lamina weight and lamina thickness had significantly lower means for kelps in the cool temperate regime compared to kelps in the warm temperate regime. When *L. pallida* morphological characteristics were grouped to wave exposure categories kruskal-wallis tests revealed significant differences amoungst categories for all morphological characteristics with lamina length, number of digits and thallus mass the exceptions. The morphological characteristics that were significantly different amoungst sites exibited similar patterns, with increasing mean values from the full sheltered to extremely sheltered categories and a decrease in mean values for the remainder of the categories. Two morphological characteristics, namely lamina thickness and stipe diameter, both exibit higher variations for the exposed categories compared to other morphological characteristics.

@@

```{r laminaria temp plots, echo=FALSE, warning=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by temperature regime.', message=FALSE, warning=FALSE, include=FALSE, fig.height=8, fig.width=7}
ggarrange(LT1, LT2,LT3,LT4,LT5,LT6,LT7,LT8,LT9, labels = c("AUTO"), common.legend = TRUE, legend = "bottom")
```

```{r lamanaria wave plots, echo=FALSE,warning=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by wave exposure category.', fig.height=8, fig.width=8}
ggarrange(LW1, LW2,LW3,LW4,LW5,LW6,LW7,LW8,LW9, labels = c("AUTO"), common.legend = TRUE, legend = "bottom")
```

Six out of the ten deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regions except for frond mass, frond length, stipe mass and epiphyte length. Mean values for primary length, primary width and stipe circumference were significantly higher for kelps from cool temperate regions than for kelps from warm temperate regions. The remaining morphological characteristics, stipe length, number of tufts and total length had significanlty lower values for kelps from cool temperate compared to kelps from the warm temperate regime. Significant differences amoungst wave exposure categories were found for all morphological characteristics of deep *E. maxima* populations. Certian morphological characteristics showed a gradual increase in value with increasing degree of wave exposure. The remaining morphological characteristics exhibited decreased values for only the higher wave exposure categories and then a sharp increase in value for the highest wave exposure category.

@@

```{r deep ecklonia temp plot, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for deep population *E. maxima* grouped by temperature.',fig.height=8, fig.width=8}
ggarrange(ET1,ET2,ET3,ET4,ET5,ET6,ET7,ET8,ET9,ET10, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

```{r deep ecklonia wave plots, warning=FALSE,fig.cap ='Boxplot of morphological characteristics for deep population *E. maxima* grouped by wave exposure.',fig.height=8, fig.width=8}
ggarrange(EW1,EW2,EW3,EW4,EW5,EW6,EW7,EW8,EW9,EW10, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

@@

```{r shallow ecklonia temp plots, echo=FALSE, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by temperature regime.',fig.height=8, fig.width=8, include=TRUE}
ggarrange(EST1,EST2,EST3,EST4,EST5,EST6,EST7,EST8,EST9,EST10, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

```{r echo=FALSE, warning=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by wave exposure.',fig.height=8, fig.width=8, include=TRUE}
ggarrange(ESW1,ESW2,ESW3,ESW4,ESW5,ESW6,ESW7,ESW8, ESW9, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

```{r rda analysis lamanaria, eval=FALSE, include=TRUE}
#data processed in the temperature and wave sections will be used here eg. abio_clim
# remove unnecessary columns
lam_df <- subset(lam_data, select = -c(1,2,6))

@@

meta_all <- read.csv("metadata/site_meta_complete.csv", sep = ";")
meta_lam_rda <- merge(lam_sites, meta_all, by = "site")
meta_lam_rda <- lam_meta_rda[-c(9), ] #remove Miller's point as it is dropping during processing

@@

#remove St Helena Bay for now as wave model not available for that site
eck_mean <- eck_mean %>% 
  filter(site != "St Helena Bay") %>% 
  filter(site != "Miller's Point")

@@

#remove date column and extra Betty's Bay row
abio_eck <- abio_eck[-c(5) , -c(2)]

@@

#summary of rda output
summary(rda_all_eck)

@@

# Parsimonious RDA
rda_eck <- rda(bio_eck ~ tp_mean + hs_mean + tp_sd + hs_sd + mean_temp, data = abio_eck)
summary(rda_eck)

@@

eck_meta_rda <- merge(eck_sites, kelp_sites, by = "site")
eck_meta_rda <- eck_meta_rda %>% 
  filter(population == "Deep")

@@

```{r rda shallow ecklonia processing, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

#filter to select only deep population ecklonia and remove Millers B and C for now...
eck_df_sh <- eck_data %>%
  filter(population == "Shallow") %>% 
  mutate(site = replace(site, site == "Buffelsbaai", "Buffelsbaai South")) %>% 
  mutate(site = replace(site, site == "Miller's Point C", "Miller's Point")) %>% 
  mutate(site = replace(site, site == "Miller's Point B", "Baboon Rock")) %>% 
  filter(site != "St James") %>% 
  filter(site != "Slangkop") %>% 
  filter(site != "Miller's Point A")

@@

#use site eck_site to align appropriate data from abio_clim data from the lamanaria processing section
abio_clim_temp <- abio_clim %>% 
  mutate(site = replace(site, site == "Sea Point", "Bakoven"))

abio_eck_sh <- merge(abio_clim, eck_sites_sh, by = "site")

@@

#summary of rda output
summary(rda_all_ecksh)

@@

step.forward <- ordistep(rda(bio_eck_sh ~ 1, data = abio_eck_sh),
                         scope=formula(rda_all_eck), direction = "forward", pstep = 1000)
#calculate variance inlfation factor
#vif.cca(rda_all_eck)

@@

# Parsimonious RDA
rda_eck_shallow <- rda(bio_eck_sh ~ hs_max + hs_mean + sd_temp + mean_temp, data = abio_eck_sh)
summary(rda_eck_shallow)

@@

# Plot RDA
rda_eck_shallow <- ggord(rda_eck_shallow,ellipse = TRUE, hull = FALSE, grp_title = "Sites", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9, coord_fix = TRUE)  # looking at the raw code, this is plotting the 'wa scores', the blue

@@

```{r rda plots, echo=FALSE, message=FALSE, warning=FALSE}
lam_rda
```

>>>>

```{r software packages and intial data prep, message=TRUE, warning=FALSE, include=FALSE, echo=FALSE}

library(tidyverse)
library(ggpubr)
library(gridExtra)
library(ggplot2)
library(ggord)
library(vegan)

meta_all <- read.csv("metadata/site_meta_complete.csv", sep = ";")

@@

```{r wave power map, eval=FALSE, fig.cap='wave power map', include=FALSE}
library(png)
library(grid)

img_wavepower <- readPNG("figures/CapePen_Total.PNG")
legend_wave_power <- readPNG("figures/CapePen_Legend.PNG")
grid.raster(img_wavepower,legend_wave_power)
```
**insert wave exposure map here**

@@

Significant differences were found between cool temperate and warm temperate regimes for all *L. pallida* morphological characteristics with lamina length, number of digits and thallus mass the exceptions. Certian morphological characteristics such as stipe mass, stipe length, total length and stipe diameter had significanlty higher means for kelps in cool temperate regimes when compared to kelps for the warm temperate regime. Conversely, lamina weight and lamina thickness had significantly lower means for kelps in the cool temperate regime compared to kelps in the warm temperate regime. When *L. pallida* morphological characteristics were grouped to wave exposure categories kruskal-wallis tests revealed significant differences amoungst categories for all morphological characteristics with lamina length, number of digits and thallus mass the exceptions. The morphological characteristics that were significantly different amoungst sites exibited similar patterns, with increasing mean values from the full sheltered to extremely sheltered categories and a decrease in mean values for the remainder of the categories. Two morphological characteristics, namely lamina thickness and stipe diameter, both exibit higher variations for the exposed categories compared to other morphological characteristics.

@@

```{r laminaria temp plots, echo=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by temperature regime.', fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
ggarrange(LT1, LT2,LT3,LT4,LT5,LT6,LT7,LT8,LT9, labels = c("AUTO"), common.legend = TRUE, legend = "bottom")
```

```{r lamanaria wave plots, echo=FALSE, fig.cap='Boxplots of *L. pallida* morphological characteristics group by wave exposure category.', fig.height=7, fig.width=7, warning=FALSE}
ggarrange(LW1, LW2,LW3,LW4,LW5,LW6,LW7,LW8,LW9, labels = c("AUTO"), common.legend = TRUE, legend = "bottom")
```

Six out of the ten deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regime except for frond mass, frond length, stipe mass and epiphyte length. Mean values for primary length, primary width and stipe circumference were significantly higher for kelps from cool temperate regions than for kelps from warm temperate regions. The remaining morphological characteristics, stipe length, number of tufts and total length had significanlty lower values for kelps from cool temperate compared to kelps from the warm temperate regime. Significant differences amoungst wave exposure categories were found for all morphological characteristics of deep *E. maxima* populations. Certian morphological characteristics showed a gradual increase in value with increasing degree of wave exposure. The remaining morphological characteristics exhibited decreased values for only the higher wave exposure categories and then a sharp increase in value for the highest wave exposure category.

@@

```{r deep ecklonia temp plot, echo=FALSE, fig.cap='Boxplot of morphological characteristics for deep population *E. maxima* grouped by temperature.', fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
ggarrange(ET1,ET2,ET3,ET4,ET5,ET6,ET7,ET8,ET9,ET10, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

```{r deep ecklonia wave plots, echo=FALSE, fig.cap='Boxplot of morphological characteristics for deep population *E. maxima* grouped by wave exposure.', fig.height=7, fig.width=7, warning=FALSE}
ggarrange(EW1,EW2,EW3,EW4,EW5,EW6,EW7,EW8,EW9,EW10, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

@@

```{r shallow ecklonia temp plots, echo=FALSE, fig.cap='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by temperature regime.', fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
ggarrange(EST1,EST2,EST3,EST4,EST5,EST6,EST7,EST8,EST9,EST10, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

```{r shallow rda wave plots, echo=FALSE, fig.cap='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by wave exposure.', fig.height=7, fig.width=7, warning=FALSE}
ggarrange(ESW1,ESW2,ESW3,ESW4,ESW5,ESW6,ESW7,ESW8, ESW9, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

```{r rda analysis lamanaria, eval=TRUE, include=FALSE, message=FALSE, warning=FALSE}
#data processed in the temperature and wave sections will be used here eg. abio_clim
# remove unnecessary columns
lam_df <- subset(lam_data, select = -c(1,2,6))

@@

meta_lam_rda <- read.csv("metadata/site_meta_complete.csv", sep = ";")
meta_lam_rda <- merge(lam_sites, meta_lam_rda, by = "site")
meta_lam_rda <- meta_lam_rda[-c(9), ] #remove Miller's point as it is dropping during processing

@@

#remove St Helena Bay for now as wave model not available for that site
eck_mean <- eck_mean %>% 
  filter(site != "Miller's Point")

@@

#remove date column and extra Betty's Bay row
abio_eck <- abio_eck[ , -c(2)]

@@

@@

# Parsimonious RDA
rda_eck <- rda(bio_eck ~ tp_mean + hs_mean + tp_sd + min_temp + dir_median, data = abio_eck)
summary(rda_eck)
anova.cca(rda_eck, step = 1000)
anova.cca(rda_eck, step = 1000, by = "axis")

@@

eck_meta_rda <- merge(eck_sites, meta_all, by = "site")
eck_meta_rda <- eck_meta_rda %>% 
  filter(population == "Deep")

@@

```{r rda shallow ecklonia processing, message=FALSE, warning=FALSE, include=FALSE}

#filter to select only deep population ecklonia and remove Millers B and C for now...
eck_df_sh <- eck_data %>%
  filter(population == "Shallow") %>% 
  mutate(site = replace(site, site == "Miller's Point A", "Miller's Point")) %>% 
  mutate(site = replace(site, site == "Bakoven", "Sea Point")) %>% 
  filter(site != "Miller's Point B") %>% 
  filter(site != "Miller's Point C") %>% 
  filter(site != "Slangkop")

@@

#use site eck_site to align appropriate data from abio_clim data from the lamanaria processing section
abio_clim_v2 <- abio_clim %>% 
  mutate(site = replace(site, site == "Bakoven", "Sea Point")) %>% 
  mutate(site = replace(site, site == "Bordjies Deep", "Bordjies"))

abio_eck_sh <- merge(abio_clim_v2, eck_sites_sh, by = "site")

@@

@@

step.forward <- ordistep(rda(bio_eck_sh ~ 1, data = abio_eck_sh),
                         scope=formula(rda_all_ecksh), direction = "forward", pstep = 1000)
#calculate variance inlfation factor
#vif.cca(rda_all_ecksh)

@@

# Parsimonious RDA
rda_eck_shallow <- rda(bio_eck_sh ~ mean_temp + dirw_median + tp_max + max_temp, data = abio_eck_sh)
summary(rda_eck_shallow)
anova.cca(rda_eck_shallow, step = 1000)
anova.cca(rda_eck_shallow, step = 1000, by = "axis")

@@

meta_all$site <- as.character(meta_all$site)
meta_shallow_rda <- meta_all %>% 
  mutate(site = replace(site, site == "Bakoven", "Sea Point")) %>%  
  mutate(site = replace(site, site == "Yzerfontein", "St Helena Bay")) %>% 
  mutate(site = replace(site, site == "Bordjies Deep", "Bordjies"))
meta_shallow_rda <- merge(meta_shallow_rda, eck_sites_sh, by = "site")

# Plot RDA
rda_eck_shallow <- ggord(rda_eck_shallow,meta_shallow_rda$region,ellipse = TRUE, hull = FALSE, grp_title = "Sites", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9, coord_fix = TRUE)  # looking at the raw code, this is plotting the 'wa scores', the blue

@@

### Abiotic drivers of *L. pallida* morphological characteristics

Forward selection, assessment of VIF and an examination of pairwise Pearson correlation coefficients allowed us to retained as the most parsimonious descriptors of *L. pallida* morphological characteristics with an adjusted *R*2 of 0.73, explaining 85% of the variation (global permutation test on final model: d.f = 4, *F* = 7.18, *p* = 0.001). The model consisted of two significant axes: RDA1 and RDA2 explained 66% and 24% of the variation, respectively. bla bla

### Abiotic drivers of *E. maxima* morphological characteristics

Forward selection, assessment of VIF and pearson correlation coefficients retained *E. maxima* morphological characteristics with an adjusted *R*2 of 0.77, explaining 92% of the variation (global permutation test on final model: d.f = 8, *F* = 6.08, *p* = 0.001). bla bla

Forward selection, assessment of VIF and pearson correlation coefficients retained *E. maxima* morphological characteristics with an adjusted *R*2 of 0.36, explaining 66% of the variation (global permutation test on final model: d.f = 3, *F* = 2.54, *p* = 0.027). bla bla bla

```{r rda plots, echo=FALSE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
ggarrange(lam_rda, rda_eck_deep, rda_eck_shallow, common.legend = TRUE)
```
